---
title: "Variance Partitioning with gamm4 Exploration"
author: "Nicholas Rachmaninoff"
date: "`r Sys.Date()`"
output: html_document
---

# Variance Partitioning: Realized vs. Raw Parameters

This document explores variance partitioning for Generalized Additive Mixed Models (GAMMs) using `gamm4`. When calculating how much variance each component (e.g., Age, Subject) explains, there are two common approaches:

1.  **Raw Variance Parameters ($\sigma^2$):** These are the estimated variances of the underlying distributions from which the random effects are drawn. While they describe the model's structure, they don't always directly reflect the variation contributed to the *specific* data points in your sample.
2.  **Realized Predictions ($var(Zu)$ and $var(X\beta)$):** This approach calculates the actual variance of the fitted values contributed by each component. 
    *   For fixed effects, we use the variance of the linear predictor: $var(X\beta)$.
    *   For random effects (including splines in `gamm4`), we use the variance of the realized predictions: $var(Zu)$.

### Why Realized Predictions are Preferred:
*   **Scale Consistency:** It puts fixed and random effects on the same scale. Fixed effects don't have a "variance parameter" in the same way random effects do, so calculating the variance of the predictions is the only way to compare them fairly.
*   **Sample Specificity:** It measures how much variation a component actually explains in your *current* dataset, rather than predicting the variation in a hypothetical infinite population.
*   **Additive Decomposition:** In many cases, the variance of the sum of these components (plus residuals) more closely approximates the total variance of the observed data, making the "Percentage Explained" much more intuitive.

# 1. Setup and Package Loading

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(gamm4)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

set.seed(123)
```

# 2. Simulate the Three Gene Scenarios

```{r simulate_data}
n_subj <- 30
n_timepoints <- 8

subjects <- data.frame(
  Subject = factor(paste0("S", sprintf("%02d", 1:n_subj))),
  Start_Age = runif(n_subj, 20, 65),
  Subj_Eff_Small = rnorm(n_subj, 0, 2),
  Subj_Eff_Large = rnorm(n_subj, 0, 15)
)

df <- expand.grid(Subject = subjects$Subject, Timepoint = 1:n_timepoints)
df <- left_join(df, subjects, by = "Subject")
df$Age <- df$Start_Age + (df$Timepoint * 1.5) + rnorm(nrow(df), 0, 0.2)

# Gene 1: Linear
df$Gene1_Linear <- 0.8 * df$Age + df$Subj_Eff_Small + rnorm(nrow(df), 0, 4)

# Gene 2: Non-Linear (Sinusoidal)
df$Gene2_NonLinear <- 15 * sin(df$Age / 6) + df$Subj_Eff_Small + rnorm(nrow(df), 0, 4)

# Gene 3: Subject Heavy
df$Gene3_SubjectHeavy <- 0.05 * df$Age + df$Subj_Eff_Large + rnorm(nrow(df), 0, 4)

df_long <- df %>%
  pivot_longer(cols = starts_with("Gene"), names_to = "Gene", values_to = "Expression") %>%
  mutate(Gene = factor(Gene, levels = c("Gene1_Linear", "Gene2_NonLinear", "Gene3_SubjectHeavy")))
```

# 3. Visualize the Simulated Data

```{r visualize_data, fig.width=10, fig.height=8}
p_age <- ggplot(df_long, aes(x = Age, y = Expression)) +
  geom_point(aes(color = Subject), alpha = 0.4) +
  geom_line(aes(group = Subject, color = Subject), alpha = 0.2) +
  geom_smooth(method = "loess", color = "black", se = FALSE, linewidth = 1.2) +
  facet_wrap(~Gene, scales = "free_y", ncol = 1) +
  theme_bw() + theme(legend.position = "none")

p_subj <- ggplot(df_long, aes(x = reorder(Subject, Expression, median), y = Expression, fill = Subject)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~Gene, scales = "free_y", ncol = 1) +
  theme_bw() + theme(legend.position = "none", axis.text.x = element_blank())

p_age | p_subj
```

# 4. Robust Variance Partition Function

```{r vp_function}
calc_vp_gamm <- function(model, spline_var = "Age", combine_age = TRUE) {
  
  # Robustly extract the lmerMod object from gamm4 result
  if (inherits(model, "gamm4")) {
    mer_obj <- model[[1]]
  } else if (is.list(model) && "mer" %in% names(model)) {
    mer_obj <- model$mer
    if (inherits(mer_obj, "gamm4")) mer_obj <- mer_obj[[1]]
  } else {
    mer_obj <- model
  }
  
  # 1. Fixed Effects Variance: var(X * beta)
  X <- getME(mer_obj, "X")
  beta <- fixef(mer_obj)
  fe_vars <- numeric()
  for (colname in colnames(X)) {
    if (colname != "X(Intercept)") {
      term_pred <- as.numeric(X[, colname, drop = FALSE] %*% beta[colname])
      fe_vars[colname] <- var(term_pred)
    }
  }
  
  # 2. Random Effects Variance: var(Z * b)
  # We use getME('Gp') to identify which columns of Z correspond to which random effect
  Z <- getME(mer_obj, "Z")
  b <- getME(mer_obj, "b")
  gp <- getME(mer_obj, "Gp")
  names_re <- names(ranef(mer_obj))
  
  re_vars <- numeric()
  for (i in seq_along(names_re)) {
    start <- gp[i] + 1
    end <- gp[i+1]
    Z_grp <- Z[, start:end, drop = FALSE]
    b_grp <- b[start:end]
    pred_grp <- as.numeric(Z_grp %*% b_grp)
    re_vars[names_re[i]] <- var(pred_grp)
  }
  
  # 3. Residuals
  var_resid <- sigma(mer_obj)^2
  
  # 4. Handle Spline Combination Logic
  # gamm4 names linear fixed effect as Xs(Age)Fx1
  spline_fixed_name <- paste0("Xs(", spline_var, ")Fx1")
  
  linear_var <- fe_vars[spline_fixed_name]
  if (is.null(linear_var) || length(linear_var) == 0 || is.na(linear_var)) linear_var <- 0
  
  nonlinear_var <- re_vars["Xr"]
  if (is.null(nonlinear_var) || length(nonlinear_var) == 0 || is.na(nonlinear_var)) nonlinear_var <- 0
  
  # Prepare final components
  # Remove the specific age parts from the general lists to avoid double counting
  fe_vars_other <- fe_vars[names(fe_vars) != spline_fixed_name]
  re_vars_other <- re_vars[names(re_vars) != "Xr"]
  
  other_fe_var <- sum(fe_vars_other)
  
  final_vars <- numeric()
  
  # Add other random effects (like Subject)
  if (length(re_vars_other) > 0) {
    for (nm in names(re_vars_other)) {
      final_vars[nm] <- re_vars_other[nm]
    }
  }
  
  if (combine_age) {
    final_vars["Age (Total)"] <- linear_var + nonlinear_var
  } else {
    final_vars["Age (Linear)"] <- linear_var
    final_vars["Age (Non-linear)"] <- nonlinear_var
  }
  
  if (other_fe_var > 0) {
    final_vars["Other (Fixed)"] <- other_fe_var
  }
  
  final_vars["Residuals"] <- var_resid
  
  # 5. Calculate Percentages
  total_var <- sum(final_vars, na.rm = TRUE)
  
  res <- data.frame(
    Component = names(final_vars),
    Variance = unname(final_vars)
  ) %>%
    mutate(Pct_Explained = (Variance / total_var) * 100) %>%
    arrange(desc(Pct_Explained))
  
  return(res)
}
```

# 5. Fit Models and Partition Variance

```{r fit_models, message=FALSE}
genes <- c("Gene1_Linear", "Gene2_NonLinear", "Gene3_SubjectHeavy")
results_list <- list()

for (g in genes) {
  # Fit Model
  fmla <- as.formula(paste(g, "~ s(Age, bs = 'cr')"))
  fit <- gamm4(fmla, random = ~(1|Subject), data = df)
  
  # Partition (Separated)
  vp_sep <- calc_vp_gamm(fit, spline_var = "Age", combine_age = FALSE)
  vp_sep$Gene <- g
  vp_sep$Type <- "Separated"
  
  # Partition (Combined)
  vp_comb <- calc_vp_gamm(fit, spline_var = "Age", combine_age = TRUE)
  vp_comb$Gene <- g
  vp_comb$Type <- "Combined"
  
  results_list[[paste0(g, "_sep")]] <- vp_sep
  results_list[[paste0(g, "_comb")]] <- vp_comb
}

# Bind all results together
df_results <- bind_rows(results_list)
```

# 6. Compare the Results

```{r compare_results, fig.width=10, fig.height=10}
# Plot 1: Separated Components
p_sep <- df_results %>%
  filter(Type == "Separated") %>%
  ggplot(aes(x = Gene, y = Pct_Explained, fill = Component)) +
  geom_bar(stat = "identity", position = "stack", color = "black", alpha = 0.9) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14) +
  labs(title = "Variance Partitioning: Separated Age Effects", y = "% Variance")

# Plot 2: Combined Components
p_comb <- df_results %>%
  filter(Type == "Combined") %>%
  ggplot(aes(x = Gene, y = Pct_Explained, fill = Component)) +
  geom_bar(stat = "identity", position = "stack", color = "black", alpha = 0.9) +
  scale_fill_manual(values = c("Age (Total)" = "#fc8d62", "Subject" = "#8da0cb", "Residuals" = "#e78ac3")) +
  theme_minimal(base_size = 14) +
  labs(title = "Variance Partitioning: Combined Age Effects", y = "% Variance")

p_sep / p_comb
```

# 7. Hypothesis Testing and Covariate Integration

A major advantage of using `gamm4` is that it provides a "dual-purpose" model. While we used the `mer` (Mixed Effects) part for variance partitioning, the `gam` (Generalized Additive Model) part is perfectly suited for hypothesis testing and p-value extraction, similar to the `dream` function in `variancePartition`.

### Why this framework is powerful:
*   **Integrated Workflow:** You fit a single model and use it for both variance partitioning (via `mer`) and inference (via `gam`).
*   **Complex Designs:** You can easily add other numeric or categorical covariates (e.g., `Sex`, `Treatment`, `BMI`) to the formula: `y ~ s(Age) + Sex + BMI`.
*   **Flexible Hypothesis Testing:** The `summary(fit$gam)` provides p-values for both the non-linear smooth terms and any linear fixed effects.

### Testing for Directionality (Increasing vs. Decreasing)
While splines can be non-monotonic, we often want to know the "overall" trend. Because `gamm4` decomposes the spline into a linear fixed effect and a non-linear random effect, we can use the sign of the linear coefficient to determine the global direction of the relationship.

```{r hypothesis_testing}
# Let's perform a test on all three simulated genes
test_genes <- c("Gene1_Linear", "Gene2_NonLinear", "Gene3_SubjectHeavy")
test_results <- list()

for (g in test_genes) {
  # Add a dummy covariate
  df$Group <- factor(rep(c("A", "B"), length.out = nrow(df)))
  
  # By using s(Subject, bs = 're') inside the formula instead of the 'random' argument,
  # the GAM summary will provide a p-value testing if the subject-level variance is non-zero.
  fmla <- as.formula(paste(g, "~ s(Age, bs = 'cr') + Group + s(Subject, bs = 're')"))
  fit <- gamm4(fmla, data = df)
  
  # # Print the summary for each gene to see the detailed stats
  # cat("\n\n--- Hypothesis Testing Summary for", g, "---\n")
  # print(summary(fit$gam))
  
  # 1. Extract P-values
  gam_sum <- summary(fit$gam)
  age_p <- gam_sum$s.table["s(Age)", "p-value"]
  subj_p <- gam_sum$s.table["s(Subject)", "p-value"]
  group_p <- gam_sum$p.table["GroupB", "Pr(>|t|)"]
  
  # 2. Determine Directionality
  lin_coef <- fixef(fit$mer)[paste0("Xs(", "Age", ")Fx1")]
  direction <- if(lin_coef > 0) "Increasing" else "Decreasing"
  
  test_results[[g]] <- data.frame(
    Gene = g,
    Age_Pval = age_p,
    Subj_Pval = subj_p,
    Group_Pval = group_p,
    Linear_Slope = lin_coef,
    Trend = direction
  )
}

# Display Table
bind_rows(test_results) %>%
  knitr::kable(digits = 4, caption = "Hypothesis Testing Results (including Subject Random Effect P-values)")
```


This approach allows you to standardize your pipeline: the same model used to calculate that "Age explains 40% of the variance" also tells you that "Age is significantly increasing with a p-value of 1e-10."

Example of gam summaries
```{r}
gam_sum

```




